# Time Use by Country Survey - Tidying and Analyzing the Data
## Data 607 - project 2 
### Heather Geiger;March 12, 2018

## About this data

This data was uploaded by Nicholas Schettini in my Data 607 class in the CUNY Master’s of Data Science program.

I have also put the raw data into my Github, available here:

https://raw.githubusercontent.com/heathergeiger/Data607_project2/master/TimeUse.csv

Nicholas gave the following description for the data:

"I found this dataset on time use by gender and by country.  Some of the variables include eating, sleeping, employment, travel, school, study, walking the dog, etc.  It seems you could analyze how males vs. females spend their time, and how each countries males and females compare to each other.  Maybe certain countries spend more time doing something more than another country; same goes for gender. "

## Loading libraries and reading in data

Load libraries.

```{r}
library(tidyr)
library(stringr)
library(dplyr)
library(ggplot2)
```

Read in data.

```{r}
timeuse <- read.csv("TimeUse.csv",header=TRUE,skipNul = TRUE,check.names=FALSE,stringsAsFactors=FALSE)
```

## Initial look at the data and initial transformations

Take a look at the file.

There are a lot of columns, so we'll display just the first 10 along with just the column names for all.

```{r, echo=TRUE,eval=FALSE}
dim(timeuse)
head(timeuse[,1:10])
colnames(timeuse)
```

```{r, echo=FALSE,eval=TRUE}
dim(timeuse)
options(width=150)
head(timeuse[,1:5])
head(timeuse[,6:7])
col8 <- data.frame(Col8 = timeuse[,8],stringsAsFactors=FALSE)
colnames(col8) <- colnames(timeuse)[8]
head(col8)
head(timeuse[,9:10])
colnames(timeuse)
```

What countries are included in this data set?

```{r}
unique(timeuse[,"GEO/ACL00"])
```

I'm assuming "Total" column will be the same for all countries, but let's check.

If so, remove this column.

Also rename "GEO/ACL00" to "Country" and "SEX" to "Sex".

```{r}
table(timeuse$Total)

timeuse <- timeuse[,setdiff(colnames(timeuse),"Total")]

colnames(timeuse)[1:2] <- c("Sex","Country")
```

Convert from wide to long format.

```{r}
dim(timeuse)
timeuse <- gather(timeuse,Activity,Time,-Sex,-Country)
dim(timeuse)

head(timeuse)
```

Write a function to convert the HH:MM notation to number of minutes.

```{r}
hours_and_minutes_to_minutes <- function(time){
	time_split <- strsplit(time,":")[[1]]
	hours <- as.numeric(time_split[1])
	minutes <- as.numeric(time_split[2])
	return((hours * 60) + minutes)
}
```

Test on a few possible options to make sure it works.

```{r}
hours_and_minutes_to_minutes("13:52")
hours_and_minutes_to_minutes("02:01")
hours_and_minutes_to_minutes("10:00")
hours_and_minutes_to_minutes("11:04")
```

Run this function on Time column.

```{r}
timeuse <- data.frame(timeuse,
	Time.in.minutes = unlist(lapply(timeuse$Time,FUN=hours_and_minutes_to_minutes)),
	stringsAsFactors=FALSE)

head(timeuse)
tail(timeuse)
```

Are there any missing values in the data?

```{r}
length(which(is.na(timeuse$Time.in.minutes) == TRUE))
```

What activities have NA for time spent?

```{r}
timeuse[is.na(timeuse$Time.in.minutes) == TRUE,]
```

These are all very specific activities that some people may have not answered for, or had on their version of the survey.

Let's change these to "00:00" and 0 minutes.

```{r}
timeuse$Time[is.na(timeuse$Time.in.minutes) == TRUE] <- "00:00"
timeuse$Time.in.minutes[is.na(timeuse$Time.in.minutes) == TRUE] <- 0
```

Let's make sure time adds up to 24 hours for all countries and genders.

```{r}
24*60
aggregate(Time.in.minutes ~ Country + Sex,FUN=sum,data=timeuse)
```

Actually, times are all over 24 hours.

Some categories must overlap.

## Resolving the over-24-hour issue

Let's pick a country and sex show all lines.

```{r}
timeuse[timeuse$Country == "Belgium" & timeuse$Sex == "Females",
c("Activity","Time","Time.in.minutes")]
```

Looks like, while the survey organizers tried their best to separate categories (e.g. "Childcare, except teaching, reading and talking" vs. "Teaching, reading and talking with child"), there is definitely some overlap.

For example, childcare could also fall under "Household and family care". And the fact that this category has a lot more time spent suggested that most people may have listed their childcare under this category instead.

I wonder if the "umbrella" categories like this are common between countries? 

We can check by getting the top say 10 activities by country and sex, and seeing which ones are repeated most often.

```{r}
timeuse <- timeuse %>% group_by(Country,Sex) %>% mutate(Activity.rank = dense_rank(-Time.in.minutes))
timeuse <- data.frame(timeuse,stringsAsFactors=FALSE)

num_country_sex_combinations_per_top10_activity <- data.frame(table(timeuse[timeuse$Activity.rank <= 10,"Activity"]))
num_country_sex_combinations_per_top10_activity$Var1 <- as.vector(num_country_sex_combinations_per_top10_activity$Var1)
num_country_sex_combinations_per_top10_activity <- num_country_sex_combinations_per_top10_activity %>% arrange(desc(Freq))

num_country_sex_combinations_per_top10_activity

timeuse[timeuse$Activity == "Travel except travel related to jobs" & timeuse$Activity.rank > 10,]
```

8 activities are found in the top 10 for all countries and sexes.

Another activity ("Travel except travel related to jobs") is found in the top 10 for all countries and sexes except French females, for whom this activity is ranked 11th.

So with one exception, 9/10 activities are all in the top 10 for all countries and sexes.

Let's now focus on how people spend their time doing these 9 activities for the remainder of the analysis.

## Additional clean-up

One additional question though - what is the deal with "Employment, related activities and travel as part of/during main and second job" vs. "Main and second job and related travel"? These sort of sound like the same thing. Let's check time spent on these by country and sex and see how they compare.

```{r}
employment_or_job <- timeuse[timeuse$Activity == "Employment, related activities and travel as part of/during main and second job" | 
	timeuse$Activity == "Main and second job and related travel",]
employment_or_job <- employment_or_job %>% select(Sex,Country,Activity,Time.in.minutes) %>% spread(Activity,Time.in.minutes)
colnames(employment_or_job)[3:4] <- c("Employment","Job")
head(employment_or_job)
ggplot(employment_or_job,aes(Employment,Job)) + 
geom_point() +
xlab("Employment, related activities and travel as part of/during main and second job") +
ylab("Main and second job and related travel") +
geom_abline(slope = 1, intercept = 0,linetype=2)
```

These are nearly identical for all combinations. I am assuming they are actually the same thing, and we should use one but not the other.

Let's use "Main and second job and related travel" but not "Employment, related activities and travel as part of/during main and second job". That will give us 8 umbrella category activities.

These categories are very broad, but collectively should give where people spend most of their time.

```{r}
top_activities <- c("Eating","Household and family care","Leisure, social and associative life","Main and second job and related travel","Personal care","Sleep","TV and video","Travel except travel related to jobs")

timeuse_top_activities <- timeuse[timeuse$Activity %in% top_activities,]
```

Now let's check the sum of time spent on all of these activities in total.

```{r}
total_hours_spent <- aggregate(Time.in.minutes ~ Country + Sex,FUN=function(x)sum(x)/60,data=timeuse_top_activities)
colnames(total_hours_spent)[3] <- "Total.hours"
head(total_hours_spent)
range(total_hours_spent$Total.hours)
```

Looks like we are still way over 24 hours.

Also there is some variation in the number of hours these activities add up to. Let's see which countries and sexes add up to more hours.

```{r}
total_hours_spent %>% spread(Sex,Total.hours) %>% arrange(Females)
total_hours_spent %>% spread(Sex,Total.hours) %>% arrange(Males)

total_hours_spent %>% spread(Sex,Total.hours) %>% mutate(Total.minutes.difference = round((Males - Females)*60))
```

Looks like France and Bulgaria are on the high side of total hours listed for these activities, for both males and females.

There are also some sex differences, which are more pronounced in some countries than others.

Let's look at one combination of sex and country again.

```{r}
timeuse_top_activities %>% filter(Sex == "Females" & Country == "Belgium") %>% arrange(Activity.rank)
```

Looking more closely, the whole data set is a bit strange.

Some of the extra hours over 24 can still be explained by umbrella categories it seems, like that maybe "Personal Care" is a superset of "Sleep" or "TV and video" is a subset of leisure.

However it seems a bit odd that no one in any of these countries has work listed as taking more than 5 hours or so of their time.

And even further removing hours that may be double-counted based on apparent umbrella categories, we are still way over on hours.

I suppose we're going to have to use the data as-is from here.

We could try to normalize by total hours, but I'm not sure how confident we are in those totals. So I think let's just compare the actual time values, with a caveat that we need to treat these comparisons with caution.

## Analysis after accepting caveats of the data

We now understand the data pretty well, including caveats we need to take when analyzing. 

Now let's make some plots!

For each of the activities, plot sets of bars by country, putting male and female side-by-side.

Start with the four activities that tended to have lower time listed, then plot for the other four.

```{r}
ggplot(timeuse_top_activities[timeuse_top_activities$Activity %in% c("Eating","Main and second job and related travel",
"Travel except travel related to jobs","TV and video"),],
aes(Country,Time.in.minutes,fill=Sex)) +
geom_bar(stat="identity",position = "dodge") +
facet_wrap(~Activity) + 
theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

Name for Germany is way too long. Let's switch to just "Germany".

Also switch United Kingdom to UK.

```{r}
timeuse_top_activities$Country <- plyr::mapvalues(timeuse_top_activities$Country,
	from = c("Germany (including  former GDR from 1991)","United Kingdom"),
	to = c("Germany","UK"))
```

```{r}
ggplot(timeuse_top_activities[timeuse_top_activities$Activity %in% c("Eating","Main and second job and related travel",
"Travel except travel related to jobs","TV and video"),],
aes(Country,Time.in.minutes,fill=Sex)) +
geom_bar(stat="identity",position = "dodge") +
facet_wrap(~Activity,scales="free_y") +
theme(axis.text.x = element_text(angle = 90, hjust = 1))

ggplot(timeuse_top_activities[!(timeuse_top_activities$Activity %in% c("Eating","Main and second job and related travel",
"Travel except travel related to jobs","TV and video")),],
aes(Country,Time.in.minutes,fill=Sex)) +
geom_bar(stat="identity",position = "dodge") +
facet_wrap(~Activity,scales="free_y") +
theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

We find that women spend a lot more time on "household and family care" according to this survey. Men spend a lot more time on "main and second job and related travel".

We also see men spending somewhat more time (though with less dramatic differences) on leisure, eating, TV and video, and non-work travel.

Amounts of time spent on personal care and sleep appear relatively similar.

Some proportion of the difference we see for men spending more time on various activities could be due to men reporting more total time spent doing different activities. But the differences we see are definitely more than the max 40 minute differences we see by sex, so this cannot explain all of what we see.

Curious how these sex differences vary by country.

For household, get ratio of female to male. For job, get ratio of male to female. Then, let's compare.

```{r}
household_and_job <- timeuse_top_activities %>% 
	filter(Activity == "Household and family care" | 
	Activity == "Main and second job and related travel") %>%
	select(Sex,Country,Activity,Time.in.minutes) %>%
	spread(Sex,Time.in.minutes) %>%
	mutate(Sex.time.ratio = ifelse(Males > Females,Males/Females,Females/Males))

ggplot(household_and_job,
aes(Country,Sex.time.ratio,fill=Activity)) +
geom_bar(stat="identity",position="dodge") +
theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
ylab("Ratio sex that spends more time/sex that spends less time")
```

Also get the time spent on these two activities separated by sex and country.

```{r}

ggplot(household_and_job %>% select(Country,Activity,Males,Females) %>% gather(Sex,Time.in.minutes,-Country,-Activity),
aes(Country,Time.in.minutes,fill=Country)) +
geom_bar(stat="identity") +
theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
facet_grid(Sex ~ Activity,scales="free_y")
```

Looks like Italy and Spain have the most extreme sex differences.

Separating each activity by sex to compare between countries, we can start to pick out what proportion of the differences might be due to women or men spending more or less time than their peers in other countries on different activities. For example, we find that Italian women are on the high end for time spent on household tasks compared to other countries, but on the low end for time spent related to a job.
